SNP Analysis

### 2019-11-18 snp/SNP_Prediction_Analysis.Rmd
checking snp data against new models

```{r setup, include=FALSE}
library(ggplot2)
library(gridExtra)
library(glmnet)
library(pander)
library(readxl)
library(pROC)
library(ggfortify)
library(preprocessCore)

source('~/workspace/shallowwgs_pipeline/lib/data_func.R')
source('~/workspace/shallowwgs_pipeline/lib/common_plots.R')

library(BarrettsProgressionRisk)

@@

adjust.cols<-function(mtx, means=NULL, sds=NULL, na.replace=0) {
  if (!is.null(means)) {
    if (ncol(mtx) != length(means))
      stop("Vector of means needs to match the columns in the matrix")
    for (i in 1:ncol(mtx)) 
      mtx[,i] = unit.var(mtx[,i], means[i], sds[i])
  } else {
    for (i in 1:ncol(mtx)) 
      mtx[,i] = unit.var(mtx[,i])
  }
  mtx[is.na(mtx)] = na.replace

  return(mtx)
}

load('~/Data/Ellie/Analysis/5e6_arms_all/model_data.Rdata', verbose = T)
load('~/Data/Ellie/Analysis/5e6_arms_all/all.pt.alpha.Rdata', verbose = T)
#alpha = '0'
alpha = '0.9'
fitV = models[[alpha]]
l = performance.at.1se[[alpha]]$lambda
features = coefs[[alpha]]

@@

pv = var(dysplasia.df[swgs_labels == 1])  # P
npv = var(dysplasia.df[swgs_labels == 0]) # NP

range(dysplasia.df[swgs_labels == 1])
range(dysplasia.df[swgs_labels == 0])

@@

qcdata = qcdata %>% rowwise %>% dplyr::mutate(
  PatientID = unlist(strsplit(Samplename, '_'))[1],
  SampleID = unlist(strsplit(Samplename, '_'))[2],
  EndoID = unlist(strsplit(Samplename, '_'))[3],
  Level = unlist(strsplit(Samplename, '_'))[4]
)
segments.list = lapply(segments.list, function(pt) {
  pt$sample = sub('\\.LogR','',pt$sample)
  pt
})

qcdata$ASCAT.SCA.ratio = apply(qcdata,1,function(s) {
  smp = subset(segments.list[[ s[['PatientID']] ]], sample == s[['Samplename']] & chr %in% c(1:22))
  smp = smp %>% rowwise %>% dplyr::mutate(
      #'Total' = nAraw + nBraw,
      'Total' = nMajor + nMinor,
      'CNV' = round(Total) - round(as.numeric(s[['Ploidy']])) )

  x = subset(smp, CNV != 0 & chr %in% c(1:22))
  sum(x$endpos - x$startpos) / chr.info[22,'genome.length']
})

@@

patient.info = as.data.frame(read_xlsx('~/Data/Ellie/Analysis/SNP/metadata_T1T2.xlsx'))
patient.info$UniqueSampleID = paste(patient.info$PatientID, patient.info$`Timepoint Code`, sep='_')
patient.info$Path.Status = patient.info$Status
patient.info[patient.info$PatientID %in% subset(patient.info, Pathology %in% c('IMC','HGD'), select='PatientID')[,1], 'Path.Status'] = 'P'

sample.list = read_xlsx('~/Data/Ellie/Analysis/SNP/20180604_Reid_1M_SampleList.xlsx', sheet = 2)
nrow(sample.list)
colnames(sample.list)[3] = 'Total.SCA'

sample.list$SCA.Ratio = sample.list$Total.SCA/max(chr.info$genome.length)

@@

sample.info = sample.info %>% rowwise() %>% dplyr::mutate(
  SCA.Ratio = ifelse(Level %in% c('BLD','GASTRIC'), 0, get.ratio(PatientID, SampleID))
)

patient.info = patient.info %>% rowwise() %>% dplyr::mutate( Endoscopy = ifelse(Timepoint == 'T1', 1, 2) )

sample.info
ggplot(sample.info, aes(ASCAT.SCA.ratio, SCA.Ratio)) + geom_point() + facet_grid(~Status)

@@

```{r, echo=F}
sm = patient.info %>% group_by(Status, Endoscopy, Pathology) %>% dplyr::tally() %>% group_by(Status, Endoscopy) %>% dplyr::mutate(freq=n/sum(n) )
sm$Pathology = factor(sm$Pathology, levels=c('NDBE','ID','LGD','HGD','IMC'), ordered = T)

@@

# Most progressors are HGD at timepoint 1
pander( round(with(subset(patient.info, Endoscopy == 1), table(Pathology, Status))['HGD',]/table(subset(patient.info, Endoscopy == 1)$Status), 3)*100, caption='% at T1')
```

@@

```{r, echo=F, warning=F, fig.height=10, fig.width=8,eval=F}

summary.file = '~/Data/Annalise/NonP/summary_logR.txt'
smy = read.table(summary.file, sep = '\t', header = T)
smy = subset(smy, Copy.number <= 12)

a = as.data.frame(smy[,c('Copy.number','var.LRR')])
colnames(a) = c('x','y')
b = as.data.frame(smy[,c('Copy.number','median.LRR')])
colnames(b) = c('x','y')
b[b$x == 2,'y'] = 0

fitSD = lm(x~y, data=a)
autoplot(fitSD)
fitM = lm(x~y, data=b)
autoplot(fitM)

fitCN = lm(y~x, data=b)

# Adjust based on the summary values per CN that we get in WGS (mostly) NP Barrett's cases
adjust.segraw<-function(segraw) {
  segraw = segraw %>% rowwise() %>% dplyr::mutate( total = nAraw+nBraw )
  segraw = subset(segraw, chr %in% c(1:22))

  segraw$adjLRR = NA
  #for (cn in unique(segraw$total)) {
  #  rows = which(segraw$total == cn)
  #  values = seg[rows, 'medLRR', drop=T]

    #newM = mean(values)
    #newSD = sd(values)

    #if (cn != 2) {
      #newM = predict(fitM, newdata=cbind.data.frame('x'=cn))
      #newSD = predict(fitSD, newdata=cbind.data.frame('x'=cn))

    #newV = mean(values)

    #}
    #segraw[rows,][['adjLRR']] = newM + (values - mean(values)) * (newSD/sd(values))
   # smy
    #median(predict(fitM, newdata=cbind.data.frame('y'=values)))-cn

    new = predict(fitCN, newdata=cbind.data.frame('x'=segraw$total))
    #plot(scale(new, center = T))

    segraw[,'adjLRR'] = scale(new,center=T)

   # segraw[rows,'adjLRR'] = (values * (predict(fitM, newdata=cbind.data.frame('y'=values)) - cn))
  #}  
segraw  
}

pt = '403'
pfiles = list.files(paste('~/Data/Reid_SNP/PerPatient', pt, sep='/'), full.names = T)
load(grep('ascat.Rdata', pfiles, value=T), verbose=F)

seg = ascat.output$segments_raw
seg = adjust.segraw(seg)

m = (melt(seg, measure.vars=c('medLRR','adjLRR')))
m$total = round(round(m$total, 1))
ggplot(m, aes(total, value, group=total)) + facet_grid(~variable) + geom_boxplot() + labs(title=pt)

seg$chr = factor(seg$chr, levels=c(1:22))
grid.arrange(
  ggplot(seg) + facet_grid(sample~chr, space='free_x', scales='free_x') +
    geom_segment(aes(x=startpos, xend=endpos, y=medLRR, yend=medLRR), color='darkgreen', size=3) + 
    theme(axis.text.x=element_blank() ) + labs(title='Low SCA, medLRR', subtitle=paste('patient',pt)),
  ggplot(seg) + facet_grid(sample~chr, space='free_x', scales='free_x') + 
    geom_segment(aes(x=startpos, xend=endpos, y=adjLRR, yend=adjLRR), color='darkred', size=3) + 
    theme(axis.text.x=element_blank() ) + labs(title='Low SCA, adjLRR', subtitle=paste('patient',pt))
)

pt = '512'
pfiles = list.files(paste('~/Data/Reid_SNP/PerPatient', pt, sep='/'), full.names = T)
load(grep('ascat.Rdata', pfiles, value=T), verbose=F)

seg = ascat.output$segments_raw
seg = adjust.segraw(seg)

m = (melt(seg, measure.vars=c('medLRR','adjLRR')))
m$total = round(round(m$total, 1))
ggplot(m, aes(total, value, group=total)) + facet_grid(~variable) + geom_boxplot() + labs(title=pt)

seg$chr = factor(seg$chr, levels=c(1:22))
grid.arrange(
  ggplot(seg) + facet_grid(sample~chr, space='free_x', scales='free_x') + 
    geom_segment(aes(x=startpos, xend=endpos, y=medLRR, yend=medLRR), color='darkgreen', size=3) + 
    theme(axis.text.x=element_blank() ) + labs(title='Mixed SCA, medLRR', subtitle=paste('patient',pt)),
  ggplot(seg) + facet_grid(sample~chr, space='free_x', scales='free_x') + 
    geom_segment(aes(x=startpos, xend=endpos, y=adjLRR, yend=adjLRR), color='darkred', size=3) + 
    theme(axis.text.x=element_blank() ) + labs(title='Mixed SCA, adjLRR', subtitle=paste('patient',pt))
)

@@

  save(mergedSegs, mergedArms, file='tmp_seg_pt.Rdata')
}
rownames(mergedSegs) = sub('\\.LogR','', rownames(mergedSegs))
rownames(mergedArms) = sub('\\.LogR','', rownames(mergedArms))

@@

copySegs = q.norm.by(mergedSegs, raw.segs)
copyArms = q.norm.by(mergedArms, raw.arms)

grid.arrange(
  ggplot( melt(raw.segs), aes(sample=value)) + stat_qq() + labs(title='Normal Q-Q plot, sWGS bins (no arms)'),
  ggplot( melt(mergedSegs), aes(sample=value)) + stat_qq() + labs(title='Normal Q-Q plot, SNP bins (no arms)'),
  ggplot( melt(copySegs), aes(sample=value)) + stat_qq() + labs(title='Normal Q-Q plot, SNP bins q-normed')
)

Adjust the variance of each genomic region by the mean and sd extracted when I trained on the sWGS data. This seems to be a more fair comparison than scaling all the new data to itself, but the predictions don't really change.

Using the mean bin values (merged during tiling).
```{r, warning=F, message=F, fig.width=10, fig.height=8}

#ns = adjust.cols(normalSegs, z.mean, z.sd)
#ns = adjust.cols(normalSegs)
#range(ns)

#na = adjust.cols(normalArms, z.arms.mean, z.arms.sd)
#na = adjust.cols(normalArms)
#range(na)

#ms = adjust.cols(mergedSegs, z.mean, z.sd)
#ms = adjust.cols(mergedSegs)
ms = adjust.cols(copySegs)
range(ms)
dim(ms)

# mergedSegs['997_122V',which(mergedSegs['997_122V',] <= 0)]
# ms['997_122V',which(mergedSegs['997_122V',] <= 0)]

#ma = adjust.cols(mergedArms, z.arms.mean, z.arms.sd)
#ma = adjust.cols(mergedArms)
ma = adjust.cols(copyArms)
range(ma)
dim(ma)

nm = melt(ms[nm.rows,])
be = melt(ms[-nm.rows,])

lims = range(nm$value, be$value, na.rm = T)
grid.arrange(
  ggplot(nm, aes(Var2, value, group=Var2, color=Var1)) + geom_point() + labs(title='Blood/gastric') + 
    theme(axis.text.x=element_blank(),legend.position='none') +ylim(lims),
  ggplot(be, aes(Var2, value, group=Var2, color=Var1)) + geom_point() + labs(title='BE samples') + 
    theme(axis.text.x=element_blank(),legend.position='none') + ylim(lims), top='Unit scaled')

@@

cx = score.cx(ms, 1)
#arrayDf = cbind(ms,ma)
arrayDf = subtract.arms(ms, ma)
#arrayDf = cbind(arrayDf, 'cx'=unit.var(cx, mn.cx, sd.cx))
arrayDf = cbind(arrayDf, 'cx'=unit.var(cx))

@@

@@

pnr = grep('BLD|gastric', preds$Samplename, ignore.case = T)
preds.normal = preds[pnr,]

@@

#randNormal = paste(unlist(strsplit(randNormal, '_'))[c(1,3)], collapse='_')
#randBE = paste(unlist(strsplit(randBE, '_'))[c(1,3)], collapse='_')

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, message = F, warning = F)
library(tidyverse)
library(gridExtra)
library(glmnet)
library(pROC)
#library(ggfortify)
library(BarrettsProgressionRisk)

#source('~/workspace/shallowwgs_pipeline/lib/data_func.R')
source('~/workspace/shallowwgs_pipeline/lib/common_plots.R')

@@

model.dir = '~/Data/Ellie/Analysis/models_5e6_all/50kb'
load(paste0(model.dir,'/model_data.Rdata'), verbose = F)
load(paste0(model.dir,'/all.pt.alpha.Rdata'), verbose = F)

alpha = '0.9'
fitV = models[[alpha]]
l = performance.at.1se[[alpha]]$lambda
features = coefs[[alpha]]

@@

@@

qcdata = qcdata %>% as_tibble %>% 
  separate(Samplename, c('PatientID','SampleID','EndoID','Level'), sep='_', remove=F)

segments.list = lapply(segments.list, function(pt) {
  pt$sample = sub('\\.LogR','',pt$sample)
  pt %>% as_tibble
})

qcdata[1,] -> s

qcdata$ASCAT.SCA.ratio = apply(qcdata,1,function(s) {
  smp = filter( segments.list[[ s$PatientID ]], sample == s$Samplename & chr %in% c(1:22) )
  smp = smp %>% #rowwise %>% 
    dplyr::mutate(
      #'Total' = nAraw + nBraw,
      'Total' = nMajor + nMinor,
      'CNV' = round(Total) - round(as.numeric(s$Ploidy)) )

  x = filter(smp, CNV != 0 & chr %in% c(1:22))
  sum(x$endpos - x$startpos)/as.numeric(chr.info[22,'genome.length'])
})

@@

patient.info = readxl::read_xlsx('~/Data/BarrettsProgressionRisk/Analysis/SNP/metadata_T1T2.xlsx') %>% 
  mutate( UniqueSampleID = paste(PatientID, `Timepoint Code`, sep='_'), Path.Status = Status) %>% 
  mutate(Path.Status = as.character(ifelse(Pathology %in% c('IMC','HGD'), 'P', Path.Status)))

sample.list = readxl::read_xlsx('~/Data/BarrettsProgressionRisk/Analysis/SNP/20180604_Reid_1M_SampleList.xlsx', sheet = 2)
colnames(sample.list)[3] = 'Total.SCA'

sample.list = sample.list %>% mutate(SCA.Ratio = Total.SCA/max(chr.info$genome.length))

@@

sample.info = sample.info %>% rowwise() %>% dplyr::mutate(
  SCA.Ratio = ifelse(Level %in% c('BLD','GASTRIC'), 0, get.ratio(PatientID, SampleID)), 
  Endoscopy = ifelse(Timepoint == 'T1', 1, 2)
)

ggplot(sample.info, aes(ASCAT.SCA.ratio, SCA.Ratio)) + geom_point() + facet_grid(~Status)

@@

```{r, echo=F}
sm = sample.info %>% group_by(Status, Endoscopy, Pathology) %>% dplyr::tally() %>% group_by(Status, Endoscopy) %>% dplyr::mutate(freq=n/sum(n) )
sm$Pathology = factor(sm$Pathology, levels=c('NDBE','ID','LGD','HGD','IMC'), ordered = T)

@@

# Most progressors are HGD at timepoint 1
pander( round(with(subset(sample.info, Endoscopy == 1), table(Pathology, Status))['HGD',]/table(subset(sample.info, Endoscopy == 1)$Status), 3)*100, caption='% HGD at T1')
```

@@

@@

  save(mergedSegs, mergedArms, file='tmp_seg_pt.Rdata')
}

rownames(mergedSegs) = sub('\\.LogR','', rownames(mergedSegs))
rownames(mergedArms) = sub('\\.LogR','', rownames(mergedArms))

@@

#copySegs = q.norm.by(mergedSegs, raw.segs)
#copyArms = q.norm.by(mergedArms, raw.arms)

training = do.call(bind_rows, purrr::map(list.files('~/Data/BarrettsProgressionRisk/Analysis/pcf_perPatient/50kb', '5e06_tiled_segvals.txt', full.names = T, recursive = T), function(f) {
  read_tsv(f, col_types = cols(.default=col_double(), 'X1'=col_character()))
}))
training.segs = as.matrix(training[,-1]); rownames(training.segs) = training$X1

training = do.call(bind_rows, purrr::map(list.files('~/Data/BarrettsProgressionRisk/Analysis/pcf_perPatient/50kb', 'arms_tiled_segvals.txt', full.names = T, recursive = T), function(f) {
  read_tsv(f, col_types = cols(.default=col_double(), 'X1'=col_character()))
}))
training.arms = as.matrix(training[,-1]); rownames(training.arms) = training$X1

merged = rbind(mergedSegs, training.segs)
merged.mean = apply(merged, 2, mean, na.rm=T)
merged.sd = apply(merged, 2, sd, na.rm=T)

for (i in 1:ncol(mergedSegs))
  mergedSegs[,i] = BarrettsProgressionRisk:::unit.var(mergedSegs[,i], merged.mean[i], merged.sd[i])

merged.arms = rbind(mergedArms, training.arms)
merged.arms.mean = apply(merged.arms, 2, mean, na.rm=T)
merged.arms.sd = apply(merged.arms, 2, sd, na.rm=T)

for (i in 1:ncol(mergedArms))
  mergedArms[,i] = BarrettsProgressionRisk:::unit.var(mergedArms[,i], merged.arms.mean[i], merged.arms.sd[i])

# grid.arrange(
#   ggplot( melt(raw.segs), aes(sample=value)) + stat_qq() + labs(title='Normal Q-Q plot, sWGS bins (no arms)'),
#   ggplot( melt(mergedSegs), aes(sample=value)) + stat_qq() + labs(title='Normal Q-Q plot, SNP bins (no arms)')
#   ggplot( melt(copySegs), aes(sample=value)) + stat_qq() + labs(title='Normal Q-Q plot, SNP bins q-normed')
# )

normal.ids = sample.info %>% filter(grepl('BLD|gastric', Level, ignore.case=T)) %>% dplyr::summarise(rows = paste0(PatientID, '_', EndoID)) %>% pull

nm.rows = which(rownames(mergedSegs) %in% normal.ids)

@@

cx = BarrettsProgressionRisk:::scoreCX(mergedSegs, 1)
#arrayDf = cbind(ms,ma)
arrayDf = BarrettsProgressionRisk:::subtractArms(mergedSegs, mergedArms)
#arrayDf = cbind(arrayDf, 'cx'=unit.var(cx, mn.cx, sd.cx))
arrayDf = cbind(arrayDf, 'cx'=BarrettsProgressionRisk:::unit.var(cx, mn.cx, sd.cx))

@@

preds = preds %>% mutate(Status = as.character(ifelse(Pathology %in% c('HGD','IMD'), 'P', Status)))

preds %>% filter(grepl('BLD|gastric',Level,ignore.case=T)) %>% group_by(Risk, Level) %>% tally %>% spread(Risk,n)

preds %>% filter(!grepl('BLD|gastric',Level,ignore.case=T)) %>% group_by(Status, Risk, Pathology) %>% tally %>% spread(Risk,n)

@@

preds.normal = preds %>% filter(grepl('BLD|gastric', Level, ignore.case=T))

preds.normal %>% group_by(Risk) %>% tally

@@

preds = sample.info %>% mutate(Sample = paste(PatientID,EndoID,sep='_')) %>% left_join(preds, by=c('Sample'='Samplename'))

summary(preds$Prob)

preds = preds %>% 
  mutate(Risk = ifelse(Prob > 0.5, 'High', 'Low')) %>% 
  mutate(Risk = factor(Risk, levels=c('Low','High'), ordered=T))

preds %>% group_by(Status, Risk) %>% tally %>% spread(Risk, n) %>%
  left_join(preds %>% group_by(Status) %>% tally, by='Status') %>%
  mutate_if(is.numeric, funs( round(./n,2))) %>% 
  dplyr::select(-n)

roc = pROC::roc(Status ~ Prob, data=preds, auc=T, ci=T, of='thresholds',transpose=T)
roc.plot(roc) + labs(title='SNP AUC by (P)')
pROC::coords(roc, 'best')

rocR = pROC::roc(Status ~ Risk, data=preds, auc=T, ci=T, of='thresholds',transpose=T)
roc.plot(rocR) + labs(title='SNP AUC by discretized risks (low/high)')
pROC::coords(rocR, 'best')
